Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-05-28 main.Rmd
minor changes to number of observations

The motivation for this work stems from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates [see @Goodwin_2012] of usage across households, such as half-hourly total usage by state, because energy companies need to plan for maximum loads on the network. But studying overall energy use hides the distributions of usage at finer scales, and makes it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data will benefit from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. Although, the motivation came through the smart meter example, this is a problem that is relevant to any temporal data observed more than once per year.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. <!--This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable.--> It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of the measured variables through these multiple observations across cyclic granularities could be a potential way to explore and bring forward different features of the  data. <!-- compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

<!-- T+1 observation not T -->
However, we need to consider the effect of number of observations on the summarization even for harmonies. Consider a data set with $(T + 1)$ observations, and two cyclic granularities $C_i$ and $C_j$ with $L$ and $M$ categories respectively <!--Each element of $C_i$ occurs approximately $(T+1)/L$ times while each element of $C_j$ occurs approximately $(T+1)/M$ times. and there are no or very few empty combinations since we only consider harmonies.--> Each combination might occur on average an equal number of times as $T\rightarrow\infty$ when the number of observations per combination is $nobs = (T + 1)/(LM)$ or they can vary anywhere between $nobs = 0$ and $nobs = (T+1)$. Now the statistical transformation used for summarizing the measured variable  for a combination should be chosen in line with this $nobs$. For example, it might be useful to compute deciles only when $nobs>=10$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

```
### Interchangeability of mappings

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (\ref{sec:summarize-measured}) or unevenly distributed events. For Kernel density based plots $nobs>30$ is suggested, whereas for summary statistics based plots like boxplots $nobs = 10$ is acceptable and for quantile plots $nobs$ should at least be equal to the number of quantiles plotted.

>>>>

The motivation for this work stems from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates [see @Goodwin_2012] of usage across households, such as half-hourly total usage by state, because energy companies need to plan for maximum loads on the network. But studying overall energy use hides the distribution of usage at finer scales, and makes it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data will benefit from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. Although, the motivation came through the smart meter example, this is a problem that is relevant to any temporal data observed more than once per year.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. <!--This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable.--> It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of the measured variable through these multiple observations could be a potential way to explore and bring forward different features of the data. <!-- compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

<!-- T+1 observation not T -->
However, we need to consider the effect of number of observations on the summarization even for harmonies. Consider a data set with $(T + 1)$ observations, and two cyclic granularities $C_i$ and $C_j$ with $L$ and $M$ categories respectively. <!--Each element of $C_i$ occurs approximately $(T+1)/L$ times while each element of $C_j$ occurs approximately $(T+1)/M$ times. and there are no or very few empty combinations since we only consider harmonies.--> Let $nobs$ be the number of observations for a combination of levels from $(C_i, C_j)$. Now, $nobs$ might be equal for all combinations on an average <!--as $T\rightarrow\infty$ when the number of observations per combination is-->  so that $nobs = (T + 1)/(LM)$ as $T\rightarrow\infty$ or ideally take any value between $\{0, 1, 2, \dots, T\}$. The statistical transformation used for summarizing the measured variable for a combination should be chosen in line with  $nobs$. For example, it might be useful to compute deciles only when $nobs \ge 10$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month are more likely to suffer from this problem.

@@

### Interchangeability of mappings

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (Section \ref{sec:summarize-measured}) or unevenly distributed events. For Kernel density based plots $nobs>30$ is suggested, for summary statistics based plots like boxplots $nobs = 10$ is acceptable and for quantile plots $nobs$ should at least be equal to the number of quantiles plotted.
